---
title: "Report for mass spectrometry data analysis pipeline"
auhtor: "`r sessionInfo()$platform`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
   rmd: "report.Rmd"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_depth: 2
    number_sections: yes
    df_print: paged
---

----------

# Background

To be added.

----------

# Prerequisites

## Packages

Loaded required R packages:

- `tidyverse`
- `ggrepel`
- `scales`
- `dendextend`
- `ggpubr`


```{r, echo = FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggrepel)
  library(scales)
  library(dendextend)
  library(ggpubr)
})
```


```{r, echo = FALSE}
# custom ggplot2 theme that is reused for all later plots
custom_colors <- c("#E7298A", "#66A61E", "#E6AB02", "#7570B3", "#B3B3B3", "#1B9E77", "#D95F02", "#A6761D")
custom_range <- function(n = 5) {
  colorRampPalette(custom_colors[c(1, 5, 2)])(n)
}

custom_theme <- function(base_size = 12, base_line_size = 1.0, base_rect_size = 1.0, ...) {
  theme_light(base_size = base_size, base_line_size = base_line_size, base_rect_size = base_rect_size) + theme(
    title = element_text(colour = grey(0.4), size = 10),
    plot.margin = unit(c(12, 12, 12, 12), "points"),
    axis.ticks.length = unit(0.2, "cm"),
    axis.ticks = element_line(colour = grey(0.4), linetype = "solid", lineend = "round"),
    axis.text.x = element_text(colour = grey(0.4), size = 10),
    axis.text.y = element_text(colour = grey(0.4), size = 10),
    panel.grid.major = element_line(linewidth = 0.6, linetype = "solid", colour = grey(0.9)),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(linetype = "solid", colour = grey(0.4), fill = NA, linewidth = 1.0),
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(colour = grey(0.4), size = 10, margin = unit(rep(3, 4), "points")),
    legend.text = element_text(colour = grey(0.4), size = 10),
    legend.title = element_blank(),
    legend.background = element_blank(),
    ...
  )
}

# set graphical parameter for subfigure labels
list_fontpars <- list(face = "plain", size = 14)

# function to export an image as svg and png
save_plot <- function(pl, path = "../figures/", width = 6, height = 6) {
  pl_name <- deparse(substitute(pl))
  svg(
    filename = paste0(path, pl_name, ".svg"),
    width = width, height = height
  )
  print(pl)
  dev.off()
  png(
    filename = paste0(path, pl_name, ".png"),
    width = width * 125, height = height * 125, res = 120
  )
  print(pl)
  invisible(capture.output(dev.off()))
}

# path to export figures
path_figures <- str_remove(snakemake@output[[1]], "report\\.html$")
```


## Result tables

Output tables from `MSstats` that were imported:

```{r, echo = FALSE}
df_tables <- snakemake@input[5:8] %>%
  unlist() %>%
  enframe(name = "table", value = "path")

for (row in 1:nrow(df_tables)) {
  assign(
    x = paste0("df_", df_tables[[row, "table"]]),
    value = read_csv(df_tables[[row, "path"]], show_col_types = FALSE)
  )
}

df_tables
```

## Sample sheet

Sample sheet listing all conditions (experimental groups) and replicates:

```{r, echo = FALSE}
df_sample_sheet <- read_tsv(
  snakemake@config$samplesheet,
  show_col_types = FALSE,
  col_names = c("file_path", "condition", "replicate", "ms_type", "control")
)
mutate(df_sample_sheet, file_path = paste0("..", str_sub(file_path, -35L, -1L)))
```

```{r, echo = FALSE}
# code chunk to set figure proportions based on N samples
figwidth <- 8.5
figheight <- 1 + round(nrow(df_sample_sheet) / 2.8, 1)
figheight2 <- 1 + 4 * ceiling(length(unique(df_comparison_result$Label)) / 3)
figheight3 <- figheight + 0.6
```

----------

# Quality control

## Summary statistics

### Number of quantified proteins

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
plot_quant_prot <- df_protein_level_data %>%
  group_by(GROUP, SUBJECT) %>%
  summarize(.groups = "drop", n_proteins = length(unique(Protein))) %>%
  ggplot(aes(x = n_proteins, y = paste(GROUP, SUBJECT))) +
  geom_col(color = "white", fill = custom_colors[1]) +
  labs(x = "", y = "", title = "Number of quantified proteins") +
  custom_theme()

# save_plot(plot_quant_prot, path = path_figures)
print(plot_quant_prot)
```

### Number of quantified peptides

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
plot_quant_pept <- df_feature_level_data %>%
  group_by(GROUP, SUBJECT) %>%
  filter(is.na(predicted)) %>%
  summarize(.groups = "drop", n_proteins = length(unique(FEATURE))) %>%
  ggplot(aes(x = n_proteins, y = paste(GROUP, SUBJECT))) +
  geom_col(color = "white", fill = custom_colors[1]) +
  labs(x = "", y = "", title = "Number of quantified peptides") +
  custom_theme()

# save_plot(plot_quant_pept, path = path_figures)
print(plot_quant_pept)
```

### Average number of features (peptides) per protein

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
plot_feat_per_prot <- df_protein_level_data %>%
  group_by(GROUP, SUBJECT) %>%
  count(NumMeasuredFeature) %>%
  ggplot(aes(x = NumMeasuredFeature, y = n)) +
  geom_step(color = custom_colors[1], linewidth = 1, direction = "vh") +
  labs(x = "", y = "", title = "Number of features (peptides) per protein") +
  facet_grid(rows = vars(paste(GROUP, SUBJECT))) +
  custom_theme()

# save_plot(plot_feat_per_prot, path = path_figures)
print(plot_feat_per_prot)
```


## Missing values

Missing values on feature level:

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight3}
plot_missing_1 <- df_feature_level_data %>%
  select(PROTEIN, GROUP, SUBJECT, predicted) %>%
  group_by(GROUP, SUBJECT) %>%
  summarize(
    .groups = "drop",
    measured = sum(is.na(predicted)),
    predicted = sum(!is.na(predicted))
  ) %>%
  pivot_longer(c("measured", "predicted"),
    names_to = "feature", values_to = "count"
  ) %>%
  mutate(feature = factor(feature, c("predicted", "measured"))) %>%
  ggplot(aes(x = count, y = paste(GROUP, SUBJECT), fill = feature)) +
  geom_col(color = "white", position = "stack") +
  labs(x = "", y = "", title = "Number of predicted and measured features") +
  custom_theme(legend.position = "bottom") +
  scale_fill_manual(values = custom_colors[c(2, 1)])

# save_plot(plot_missing_1, path = path_figures)
print(plot_missing_1)
```

### Missing values on protein level

This plot shows percent (%) feature inference that was used to quantify each individual protein.
None-imputed (fully quantified) proteins were removed to make the figure appear more clean.

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
plot_missing_2 <- df_protein_level_data %>%
  select(Protein, GROUP, SUBJECT, MissingPercentage) %>%
  filter(MissingPercentage > 0) %>%
  ggplot(aes(x = MissingPercentage * 100, y = paste(GROUP, SUBJECT))) +
  geom_violin(color = "white", fill = custom_colors[1], alpha = 0.3) +
  geom_jitter(height = 0.2, color = custom_colors[1]) +
  labs(
    x = "", y = "", title = "Percent imputed features per protein"
  ) +
  custom_theme()

# save_plot(plot_missing_2, path = path_figures)
print(plot_missing_2)
```

### Number of proteins quantified in number of samples

Stepplot showing number (N) of proteins quantified in number of samples.
Large area under the curve means that many proteins have been quantified in a majority of the samples.

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
plot_stepplot <- df_protein_level_data %>%
  group_by(Protein) %>%
  summarize(n_samples = sum(!is.na(LogIntensities))) %>%
  count(n_samples, name = "n_proteins") %>%
  arrange(desc(n_samples)) %>%
  add_row(n_samples = nrow(df_sample_sheet), n_proteins = 0, .before = 0) %>%
  add_row(n_samples = 0, n_proteins = 0) %>%
  mutate(n_proteins = cumsum(n_proteins)) %>%
  ggplot(aes(x = n_proteins, y = n_samples)) +
  geom_step(color = custom_colors[1], linewidth = 1, direction = "vh") +
  labs(
    x = "N proteins", y = "N samples",
    title = "N proteins quantified in N samples"
  ) +
  custom_theme()

# save_plot(plot_stepplot, path = path_figures)
print(plot_stepplot)
```


### Problematic protein quantifications

Protein quantifications where more than 50% of _feature_ quantifications are missing.

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight}
plot_problematic <- df_protein_level_data %>%
  group_by(GROUP, SUBJECT) %>%
  count(more50missing) %>%
  filter(more50missing) %>%
  ggplot(aes(x = n, y = paste(GROUP, SUBJECT))) +
  geom_col(color = "white", fill = custom_colors[1]) +
  labs(x = "", y = "", title = "Number of proteins with > 50% missing features") +
  custom_theme()

# save_plot(plot_problematic, path = path_figures)
print(plot_problematic)
```

----------

# Protein quantification

## Sample-wise comparison

### Normalized log2 protein abundance

Protein abundance as calculated by MSstats after normalization. The default is per-group median normalization.
Normalization options can be changed in the `config.yml` that was specified when ececuting the pipeline.
Color indicares the number of the individual replicates.

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight3}
plot_log2_abundance <- df_protein_level_data %>%
  ggplot(aes(x = LogIntensities, y = GROUP, color = factor(SUBJECT))) +
  geom_vline(xintercept = 0, color = grey(0.75), linetype = 2) +
  geom_boxplot(outlier.shape = NA) +
  labs(x = "", y = "", title = "Normalized log2 protein abundance") +
  custom_theme(legend.position = "bottom") +
  scale_color_manual(values = custom_colors)

# save_plot(plot_log2_abundance, path = path_figures)
print(plot_log2_abundance)
```

### Profile plots

Comparison of feature intensity profile for selected protein(s). 
This plot can be used to evaluate the protein quantification from feature intensity.
This is a `ggplot` adaption of the default profile plot output from MSstats.
A selection of 10 proteins with different total abundance is chosen as representative sample.
Each of these proteins represents the middle value of one of 10 abundance ranges.

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = 12}
list_diff_abund_prot <- df_protein_level_data %>%
  group_by(Protein) %>%
  summarize(median_int = median(LogIntensities, na.rm = TRUE)) %>%
  mutate(int_group = cut_interval(median_int, n = 10)) %>%
  group_by(int_group) %>%
  arrange(median_int) %>%
  slice(round(n() / 2)) %>%
  pull(Protein)

plot_profile <- ggplot() +
  geom_line(
    data = df_feature_level_data %>%
      filter(PROTEIN %in% list_diff_abund_prot),
    mapping = aes(x = factor(RUN), y = ABUNDANCE, group = PEPTIDE),
    color = grey(0.8)
  ) +
  geom_point(
    data = df_protein_level_data %>%
      rename(PROTEIN = Protein) %>%
      filter(PROTEIN %in% list_diff_abund_prot),
    mapping = aes(x = factor(RUN), y = LogIntensities, color = GROUP),
    size = 2
  ) +
  labs(
    x = "", y = expression("log"[2] * " intensity"),
    title = "Feature intensity (grey) and inferred protein quantity (color)"
  ) +
  custom_theme(legend.position = "bottom") +
  scale_color_manual(values = custom_colors) +
  facet_grid(rows = vars(factor(PROTEIN, list_diff_abund_prot)))

# save_plot(plot_profile, path = path_figures, width = 8, height = 12)
print(plot_profile)
```

### Similarity of conditions and replicates

To visualize similarity between samples, principal component analysis is used. 
Conditions are encoded by color and replicates are indicated with numbers.

```{r, echo = FALSE, fig.width = 7, fig.height = 7, warning = FALSE}
pca_result <- df_protein_level_data %>%
  select(Protein, GROUP, SUBJECT, LogIntensities) %>%
  pivot_wider(names_from = c("GROUP", "SUBJECT"), values_from = "LogIntensities") %>%
  column_to_rownames("Protein") %>%
  as.matrix() %>%
  t() %>%
  replace(., is.na(.), 0) %>%
  prcomp()

df_PCA <- pca_result$x %>%
  as_tibble(rownames = "sample") %>%
  mutate(
    condition = str_sub(sample, 1, -3),
    replicate = str_sub(sample, -1, -1)
  )

plot_pca <- df_PCA %>%
  ggplot(aes(x = PC1, y = -PC2, size = PC3, color = condition, label = replicate)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(size = 4, show.legend = FALSE) +
  labs(
    title = "PCA, first three principal components",
    subtitle = "Point size encodes PC3", x = "PC1", y = "PC2"
  ) +
  custom_theme(legend.position = "bottom", aspect = 1) +
  scale_color_manual(values = custom_colors[1:8]) +
  guides(size = "none")

# save_plot(plot_pca, path = path_figures, width = 7, height = 7)
print(plot_pca)
```

## Group-wise comparison

### Model QC

Internally, MSstats fits statistical models to protein quantifications for each comparison, similar to mixed linear models.
Such models can be fitted for multiple comparisons between conditions, particularly where multi-factorial data sets are involved
(that means, conditions characterized by different variables simultaneously).

To check if the assumptions of the model fitting are met, a [QQ (quantile quantile) plot](https://en.wikipedia.org/wiki/Q%E2%80%93Q_plot) 
can be used. The most important assumption is the normal distribution of measurements and errors. Errors should scale with the measurement,
and QQ plots test this assumption by plotting the theoretical quantiles against the observed quantiles of the measurements. 
The expected behavior is that the plots follow a line. If not, protein quantifications are not normal-distributed.
The follwing figure shows QQ plots for the 10 exemplary proteins from the previous section.

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = 12}
plot_qq <- df_model_qc %>%
  filter(Protein %in% list_diff_abund_prot) %>%
  ggplot(aes(sample = ABUNDANCE, color = GROUP, group = NA)) +
  stat_qq(color = grey(0.5)) +
  stat_qq_line() +
  labs(
    title = "QQ plot, theoretical vs. observed quantiles of feature abundance",
    x = "theoretical quantiles", y = "variable"
  ) +
  custom_theme(aspect = 1) +
  facet_wrap(~Protein, ncol = 3, nrow = 4, scales = "free")

# save_plot(plot_qq, path = path_figures, width = 9, height = 12)
print(plot_qq)
```

### Volcano plot

Volcano plots are useful to evaluate the global changes between conditions. They show the log2 fold change (`log2FC`)
on the x-axis and the p-value on the y-axis. Both metrics are computed by MSstats for the comparisons ('constrasts')
indicated in the sample sheet. The p-value is the Benjamini-Hochberg adjusted p-value from MSstats default statistical test.

Significance boundaries are shown that indicate which genes fall within a range that is both significant in terms of
p-value, and effect size (log2 fold change). The standard boundaries for both metrics are:

- alpha significance threshold = 0.01
- fold change threshold = 2

The boundary represents the _product_ of the fold change and -log10 adjusted p-value thresholds.
Example: All points on this line are equivalent to _k_ = -log10(0.01) * 2 = 4.

```{r, echo = FALSE, warnings = FALSE}
threshold_pvalue <- 0.01
threshold_fc <- 2
```

```{r, echo = FALSE, warning = FALSE, fig.width = figwidth, fig.height = figheight2}
geom_effect_boundaries <- function(alpha, fc, xlim = c(-8, 8), color = grey(0.3)) {
  x <- c(seq(xlim[1], 0, 0.1), seq(0, xlim[2], 0.1))
  y <- (-log10(alpha) * fc) / c(seq(xlim[2], 0, -0.1), seq(0, xlim[2], 0.1))
  df <- data.frame(x = x, y = y)
  geom_line(data = df, aes(
    x = x, y = y,
    shape = NULL, col = NULL
  ), lty = 2, color = color)
}

plot_volcano <- df_comparison_result %>%
  ggplot(aes(
    x = log2FC, y = -log10(adj.pvalue),
    color = abs(log2FC) * -log10(adj.pvalue) >=
      threshold_fc * -log10(threshold_pvalue)
  )) +
  geom_point(size = 0.5) +
  geom_effect_boundaries(
    alpha = threshold_pvalue,
    fc = threshold_fc,
    xlim = c(-15, 15)
  ) +
  labs(
    title = "Volcano plot",
    subtitle = paste0(
      "color indicates proteins with score equivalent to log2FC > ",
      threshold_fc, " and p-value < ", threshold_pvalue
    ),
    x = expression("log"[2] * " FC"), y = expression("-log"[10] * " p-value")
  ) +
  lims(x = c(-7, 7), y = c(-0.1, 8.1)) +
  custom_theme(legend.position = "bottom", aspect = 1) +
  scale_color_manual(values = custom_colors[c(5, 1)]) +
  facet_wrap(~Label,
    ncol = 3,
    nrow = ceiling(length(unique(df_comparison_result$Label)) / 3)
  )

# save_plot(plot_volcano, path = path_figures)
print(plot_volcano)
```

### List of differentially abundant proteins

This table lists all proteins that have been marked with colored points in the volcano plot.
THe table is in _long format_, that means quantifications are shown in one column and descriptors (e.g. comparisons) in another.

```{r, echo = FALSE, warnings = FALSE}
df_significant <- df_comparison_result %>%
  group_by(Protein) %>%
  filter(
    !is.na(adj.pvalue),
    any(
      abs(log2FC) >= log2(threshold_fc) &
        adj.pvalue <= threshold_pvalue
    )
  ) %>%
  ungroup()

if (nrow(df_significant)) {
  print(paste0(
    "A total of ", length(unique(df_significant$Protein)),
    " proteins were found with at least one significant change in a comparison"
  ))
  df_significant
} else {
  print("No significantly changed proteins have been found with the default thresholds.")
}
```

### Heat map

The heat map shows a direct comparison of protein abundance for all conditions, including only the proteins falling below
the previously defined thresholds.

```{r, echo = FALSE, warnings = FALSE}
# this function allows
# originally contained in this R package: https://github.com/m-jahn/R-tools
fct_cluster <- function(variable, group, value, method = "ward.D2") {
  variable <- as.character(variable)
  df <- tidyr::tibble(variable = variable, group = group, value = value)
  df <- tidyr::complete(df, variable, group)
  df <- tidyr::pivot_wider(df,
    names_from = group, values_from = value,
    values_fn = function(x) {
      replace(x, is.infinite(x), NA)
    }
  )
  mat <- as.matrix(tibble::column_to_rownames(df, var = "variable"))
  dist_mat <- replace(dist(mat), is.na(dist(mat)), 0)
  cl <- stats::hclust(dist_mat, method = "ward.D2")
  ord <- stats::order.dendrogram(stats::as.dendrogram(cl))
  factor(variable, rownames(mat)[ord])
}
```


```{r, echo = FALSE, warnings = FALSE, fig.width = 6, fig.height = 8}
if (nrow(df_significant)) {
  # reduce number of proteins before plotting heat map
  if (length(unique(df_significant$Protein)) > 50) {
    df_significant <- df_significant %>%
      group_by(Label) %>%
      arrange(desc(abs(log2FC))) %>%
      slice(1:50)
    subtitle_reduced <- "Number of proteins was reduced to 50."
  } else {
    subtitle_reduced <- ""
  }
  df_significant %>%
    mutate(Protein = fct_cluster(Protein, Label, log2FC)) %>%
    ggplot(aes(
      y = Protein,
      x = Label, fill = log2FC
    )) +
    geom_tile() +
    custom_theme(legend.pos = "right") +
    labs(
      title = paste0("Heat map of protein abundance (n = ", length(unique(df_significant$Protein)), ")"),
      subtitle = subtitle_reduced,
      x = "",
      y = ""
    ) +
    theme(
      axis.text.x = element_text(angle = 90),
      axis.ticks.x = element_blank()
    ) +
    scale_fill_gradientn(
      colours = c(custom_colors[1], grey(0.9), custom_colors[2]),
      limits = c(-6, 6)
    )

  # save_plot(plot_heatmap, path = path_figures)
  print(plot_heatmap)
} else {
  print("No significantly changed proteins have been found with the default thresholds.")
}
```

----------

# About this report

## Data accessability

The following resources were used to generate this report:

- MSstats result tables: `r paste0(snakemake@config$output, "msstats/")`.
- sample sheet ('manifest' in fragpipe language): `r snakemake@config$samplesheet`
- fragpipe workflow: `r snakemake@config$workflow`
- organism database: `r snakemake@config$database`
- fragpipe binary: `r snakemake@config$fragpipe$path`

## Session Info

Link to the R markdown source that was used to generate this report:

<a download="report.Rmd" href="`r base64enc::dataURI(file = params$rmd, mime = 'text/rmd', encoding = 'base64')`">R Markdown source file</a>

Session info (R base and package versions):

```{r, echo = FALSE}
# save.image("/home/michael/Desktop/rws.Rdata")
sessionInfo()
```

